Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-12-03 manuscript/manuscript.Rmd
closes #3

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use.  However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and advanced understanding required to interpret results.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper, we use this association to provide estimates of chlorophyll *a* concentrations that if exceeded would be indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recreational use.  For the specific advisory microcystin concentrations of 0.3, 1, 1.6, and 2, we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microcystin levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

@@

Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and well understood components of water quality that are readily assessed in the field.  For instance, there are small or hand held fluorometers that measure chlorohpyll *a*.  Additionally, chlorophyll *a* is a very commonly measured component of water quality that is also known to be positively associated with Microsystin-LR concentrations [@pip2014microcystin; @yuan2014managing].  Yuan et. al [-@yuan2014managing] explore these associations in detail and control for other related variables.  In their analysis they find that total nitrogen and chlorophyll *a* show the strongest association with microcystin. Furthermore, they identify chlorophyll*a* and total nitrogen concentrations that are associated with exceeding 1 $\mu$g/L of microcystin.   Given these facts, it should be possible to identify chlorophyll *a* concentrations that would be associated with the new USEPA Microcystin-LR health advisory levels for drinking water.  Identifying these associations would provide another tool for water resource managers to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisory changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.

@@

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`%, `r fn_percent$who_drink*100`%, `r fn_percent$epa_adult*100`% and , `r fn_percent$who_rec*100`% for the U.S EPA childrens drinking water, the WHO drinking water, the USEPA adult drinking water, and the WHO recreational advisories, respecitvely.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.

@@

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous meausrement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases.

>>>>

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use.  However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and advanced understanding required to interpret results.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  In particular, chlorophyll *a* is very commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper, we use this association to provide estimates of chlorophyll *a* concentrations that if exceeded would be indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA health advisories for drinking water as well as the World Health Organization levels for drinking water and recreational use.  For the specific advisory microcystin concentrations of 0.3, 1, 1.6, and 2, we find chlorophyll *a* concentrations of `r mc_chla[1]`, `r mc_chla[2]`, `r mc_chla[3]`, and `r mc_chla[4]`, respectively.  When managing for these various microcystin levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

@@

Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and well understood components of water quality that are readily assessed in the field.  For instance, there are small or hand held fluorometers that measure chlorohpyll *a*.  Additionally, chlorophyll *a* is a very commonly measured component of water quality that is also known to be positively associated with microsystin-LR concentrations [@pip2014microcystin; @yuan2014managing].  Yuan et. al [-@yuan2014managing] explore these associations in detail and control for other related variables.  In their analysis they find that total nitrogen and chlorophyll *a* show the strongest association with microcystin. Furthermore, they identify chlorophyll*a* and total nitrogen concentrations that are associated with exceeding 1 $\mu$g/L of microcystin.   Given these facts, it should be possible to identify chlorophyll *a* concentrations that would be associated with the new USEPA Microcystin-LR health advisory levels for drinking water.  Identifying these associations would provide another tool for water resource managers to help manage the threat to public health posed by cHABs and would be especially useful in the absence of microcystin-LR concentrations.

@@

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisories changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also caclulated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.

@@

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of `r fn_percent$epa_child*100`%, `r fn_percent$who_drink*100`%, `r fn_percent$epa_adult*100`% and , `r fn_percent$who_rec*100`% for the U.S EPA childrens drinking water, the WHO drinking water, the USEPA adult drinking water, and the WHO recreational advisories, respectively.  In each case we miss less than 10% of the lakes that are in fact exceeding the microcystin-LR advisory.

@@

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous measurement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases.

### 2015-12-03 manuscript/manuscript.tex
closes #3

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper, we use this association to provide estimates
of chlorophyll \emph{a} concentrations that if exceeded would be
indicative of a higher probability of exceeding select health advisory
concentrations for microcystin-LR. Using the 2007 National Lakes
Assessment and a conditional probability approach, we identify
chlorophyll \emph{a} concentrations that are more likely than not to be
associated with an exceedance of a microcystin health advisory level. We
look at the recent US EPA standards for drinking water as well as the
World Health Organization levels for drinking water and recreational
use. For the specific advisory microcystin concentrations of 0.3, 1,
1.6, and 2, we find chlorophyll \emph{a} concentrations of 0.07, 0.07,
2.79, and 11.36, respectively. When managing for these various
microcystin levels, exceeding these reported chlorophyll \emph{a}
concentrations should be a trigger for further testing and possible
management action.

@@

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms (Rinta-Kanto et
al. 2009, HABHRCA 2014, Jetoo et al. 2015), and in response the US
Environmental Protection Agency (USEPA) has recently released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005, USEPA 2015). Additionally, the World Health Organization (WHO) has
had proposed advisory levels for drinking water and a range of
recreational risk levels (Chorus and Bartram 1999, (WHO) and others
2003). While these levels and associated advisories are likely to help
mitigate the impacts from harmful algal blooms, they are not without
complications.

@@

Fortunately, microcystin-LR has been shown to be associated with several
other, more commonly measured and well understood components of water
quality that are readily assessed in the field. For instance, there are
small or hand held fluorometers that measure chlorohpyll \emph{a}.
Additionally, chlorophyll \emph{a} is a very commonly measured component
of water quality that is also known to be positively associated with
Microsystin-LR concentrations (Pip and Bowman 2014, Yuan et al. 2014).
Yuan et. al (2014) explore these associations in detail and control for
other related variables. In their analysis they find that total nitrogen
and chlorophyll \emph{a} show the strongest association with
microcystin. Furthermore, they identify chlorophyll\emph{a} and total
nitrogen concentrations that are associated with exceeding 1 \(\mu\)g/L
of microcystin. Given these facts, it should be possible to identify
chlorophyll \emph{a} concentrations that would be associated with the
new USEPA Microcystin-LR health advisory levels for drinking water.
Identifying these associations would provide another tool for water
resource managers to help manage the threat to public health posed by
cHABs and would be especially useful in the absence of microcystin-LR
concentrations.

@@

Conditional probability analysis provides information about the
probability of observing one event given another event has also occured.
For this analysis, we used CPA to examine how the conditional
probability of exceeding one of the health advisory changes as
chlorophyll \emph{a} increases in a lake. We expect to find higher
chlorohpyll \emph{a} concentrations to be associated with higher
probabilities of exceeding the microcystin-LR health advisory levels. We
also caclulated bootstrapped 95\% confidence intervals (CI) using 1000
bootstrapped samples. Thus, to identify chlorophyll \emph{a}
concentrations of concern we identfiy the value of the upper 95\% CI
across a range of conditional probabilities of exceeding each health
advisory level. As both microcystin-LR and chlorophyll \emph{a} values
were highely skewed right, a log base 10 transformation was used.
Additional details of the specific implementation are available at
\url{https://github.com/USEPA/microcystinchla}. A more detailed
discussion of CPA is beyond the scope of this paper, but see Paul et al.
(2005) and Hollister et al. (2008) for greater detail.

@@

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 0\%, 0\%, 0\% and , 1\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respecitvely. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory.

@@

Lastly, using chlorophyll \emph{a} is not meant as a replacement for
testing of microcystin-LR. It should be used when other, direct
measurements of cyanotoxins are not available. In those cases, which are
likely to be common at least in the near future, using a more ubiquitous
meausrement, such as chlorophyll \emph{a} will provide a reasonable
proxy for the probability of exceeding a microcystin-LR health advisory
level and provide better protection against adverse effects in both
drinking and recreational use cases.

@@

@@

(WHO), W. H. O., and others. 2003. Cyanobacterial toxins: Microcystin-lR
in drinking-water. background document for development of wHO guidelines
for drinking-water quality. geneva, switzerland. World Health
Organization, 2nd ed. Geneva.

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper, we use this association to provide estimates
of chlorophyll \emph{a} concentrations that if exceeded would be
indicative of a higher probability of exceeding select health advisory
concentrations for microcystin-LR. Using the 2007 National Lakes
Assessment and a conditional probability approach, we identify
chlorophyll \emph{a} concentrations that are more likely than not to be
associated with an exceedance of a microcystin health advisory level. We
look at the recent US EPA health advisories for drinking water as well
as the World Health Organization levels for drinking water and
recreational use. For the specific advisory microcystin concentrations
of 0.3, 1, 1.6, and 2, we find chlorophyll \emph{a} concentrations of
0.07, 0.07, 2.79, and 11.36, respectively. When managing for these
various microcystin levels, exceeding these reported chlorophyll
\emph{a} concentrations should be a trigger for further testing and
possible management action.

@@

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms (Rinta-Kanto et
al. 2009, HABHRCA 2014, Jetoo et al. 2015), and in response the US
Environmental Protection Agency (USEPA) has recently released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005, USEPA 2015). Additionally, the World Health Organization (WHO) has
had proposed advisory levels for drinking water and a range of
recreational risk levels (Chorus and Bartram 1999, Organization 2003).
While these levels and associated advisories are likely to help mitigate
the impacts from harmful algal blooms, they are not without
complications.

@@

Fortunately, microcystin-LR has been shown to be associated with several
other, more commonly measured and well understood components of water
quality that are readily assessed in the field. For instance, there are
small or hand held fluorometers that measure chlorohpyll \emph{a}.
Additionally, chlorophyll \emph{a} is a very commonly measured component
of water quality that is also known to be positively associated with
microsystin-LR concentrations (Pip and Bowman 2014, Yuan et al. 2014).
Yuan et. al (2014) explore these associations in detail and control for
other related variables. In their analysis they find that total nitrogen
and chlorophyll \emph{a} show the strongest association with
microcystin. Furthermore, they identify chlorophyll\emph{a} and total
nitrogen concentrations that are associated with exceeding 1 \(\mu\)g/L
of microcystin. Given these facts, it should be possible to identify
chlorophyll \emph{a} concentrations that would be associated with the
new USEPA Microcystin-LR health advisory levels for drinking water.
Identifying these associations would provide another tool for water
resource managers to help manage the threat to public health posed by
cHABs and would be especially useful in the absence of microcystin-LR
concentrations.

@@

Conditional probability analysis provides information about the
probability of observing one event given another event has also occured.
For this analysis, we used CPA to examine how the conditional
probability of exceeding one of the health advisories changes as
chlorophyll \emph{a} increases in a lake. We expect to find higher
chlorohpyll \emph{a} concentrations to be associated with higher
probabilities of exceeding the microcystin-LR health advisory levels. We
also caclulated bootstrapped 95\% confidence intervals (CI) using 1000
bootstrapped samples. Thus, to identify chlorophyll \emph{a}
concentrations of concern we identfiy the value of the upper 95\% CI
across a range of conditional probabilities of exceeding each health
advisory level. As both microcystin-LR and chlorophyll \emph{a} values
were highely skewed right, a log base 10 transformation was used.
Additional details of the specific implementation are available at
\url{https://github.com/USEPA/microcystinchla}. A more detailed
discussion of CPA is beyond the scope of this paper, but see Paul et al.
(2005) and Hollister et al. (2008) for greater detail.

@@

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 0\%, 0\%, 0\% and , 1\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respectively. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory.

@@

Lastly, using chlorophyll \emph{a} is not meant as a replacement for
testing of microcystin-LR. It should be used when other, direct
measurements of cyanotoxins are not available. In those cases, which are
likely to be common at least in the near future, using a more ubiquitous
measurement, such as chlorophyll \emph{a} will provide a reasonable
proxy for the probability of exceeding a microcystin-LR health advisory
level and provide better protection against adverse effects in both
drinking and recreational use cases.

@@

Organization, W. H. 2003. Cyanobacterial toxins: Microcystin-lR in
drinking-water. background document for development of wHO guidelines
for drinking-water quality. geneva, switzerland. World Health
Organization, 2nd ed. Geneva.

@@
